Location and Scale-Invariant Power Transformations for Transforming Data to

### 2024-09-17 manuscript/manuscript.Rmd
Revised discussion.

We used simulated data to assess invariance to location and scale of the proposed power transformations, weighting for robust transformations, and to develop the empirical central normality test.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well as $\lambda$, $x_0$ and $s$ parameters for location- and scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.
The required algorithms were implemented in the `power.transform` R software package [@Zwanenburg2024-kq].
Of note, the `power.transform` package shifts feature values into the positive domain if negative or zero values are present for Box-Cox power transformations.

@@

Location- and scale-invariant transformations are realised by simultaneously optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
Alternatively, standardisation of a numeric variable (e.g., through subtracting its median value and division by its interquartile range) prior to conventional power transformations may achieve a similar effect in reducing sensitivity to the distribution's location and scale.

We assessed several methods for robust power transformation.
Methods using the z-score of the transformed feature or the residual error yielded worse results than the non-robust method.
Different initial values, closer to the upper limits ($\delta_1 = 10.0; \delta_2 = 10.0$), led to a reduced loss.
However, these values effectively correspond to a non-robust transformation, where all instances receive the same weight.
Underperformance of these methods could be explained by their use of transformed features for setting weights.
In essence, parameter optimisation with MLE also affects weights.
This may lead to different optimal values for transformation parameter $\lambda$, which increases the presented loss.

The empirical test for central normality differs from other tests for normality in that the test statistic is independent of the number of samples.
This makes this test more practical for assessing central normality for larger sample numbers, where other test may detect trivial deviations from normality.

As a limitation, we did observe several numerical issues for optimisation criteria other than MLE (appendix B).

>>>>

We used simulated data to assess invariance to location and scale of the proposed power transformations, weighting for robust transformations, and to develop the empirical central normality test.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well as $\lambda$, $x_0$ and $s$ parameters for location- and scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.
The required algorithms were implemented in the `power.transform` R software package [@Zwanenburg2024-kq] (version 1.0.0).
Of note, the `power.transform` package shifts feature values into the positive domain if negative or zero values are present for Box-Cox power transformations.

@@

Location- and scale-invariant transformations are realised by simultaneously optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
We derived the log-likelihood function to facilitate optimisation using MLE.
Alternatively, standardisation of a numeric variable (e.g., through subtracting its median value and division by its interquartile range) prior to conventional power transformations may achieve a similar effect in reducing sensitivity to the distribution's location and scale.
While this alternative helps prevent these issues -- provided that normalisation does not lead to negative values for Box-Cox transformation -- location- and scale-invariant transformations seem to provide an overall better transformation to normality (Appendix F).

We assessed several methods for robust power transformation.
Methods that relied on the z-score of the transformed feature or the residual error yielded worse results than the non-robust method.
This is partly due to the initial choice of threshold parameters.
Using different initial values, closer to the upper limits ($\delta_1 = 10.0; \delta_2 = 10.0$), led to a reduced loss.
However, these values effectively correspond to a non-robust transformation, where all instances receive the same weight.
Underperformance of these methods could be explained by their use of transformed features for setting weights.
Consequently, the weights change at each iteration in the MLE optimisation process.
This increases local variance in the log-likelihood function and create local optima that the optimiser may not handle well.
As a consequence, optimal values for transformation parameter $\lambda$ might differ, which increases the presented loss for optimising threshold parameters.
Methods that relied on the empirical probability did not suffer from this issue, as weights remained fixed during MLE.

We introduced an empirical test for central normality to assess whether distributions deviate from normality in a way that might require closer inspection prior to further processing.
The empirical test for central normality differs from other tests for normality, such as the Shapiro-Wilk test [@Shapiro1965-zd], in that the test statistic is independent of the number of samples.
This makes this test more practical for assessing central normality for larger sample numbers, where other test may detect inconsequential deviations from normality.

This work has the following limitation.
Firstly, we did observe several numerical stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to very large or small numbers when using conventional power transformations.
For MLE stability issues were not observed.
Secondly, the empirical central normality test is based on simulations instead of statistical theory, and relies on a somewhat arbitrary definition of the central portion of a distribution.
Thus, while the test may asses whether data is sufficiently normally distributed for practical purposes, it should not be used as a strict test for normality.
